Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-12 analysis/paper/paper.Rmd
incorporated KS and RC edits and re-knit paper.docx

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and consequently have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume \~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Chinook Salmon in the Upper Salmon River are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for Chinook Salmon in the Upper Salmon River as it supports rearing and migration of juveniles from all populations including fall DSR and spring NRR migrants.

@@

Next, our abundance estimates assumed that capture and recapture events are random samples of the population. Angling methods often have a size selection bias, thereby limiting abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. The estimators also assumed equal catchability of individual fish between sampling events; two fish were recaptured during multiple days within a survey, indicating that marked fish were continually susceptible to angling. It is also possible that marked and released fish may not have immediately mixed thoroughly back into the population. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may have led to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. As an example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate. This would additionally account for marked and released fish being "hook shy" for a period of time. The adjusted delayed-mixing Schnabel estimator may account for these biases; however the magnitude of bias is unknown.

@@

Our bioenergetics approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999]. Other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration, and so generalizing energy densities among salmonids may not be problematic. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during fall months when DSR Chinook Salmon are the dominant prey species available the Salmon River and Deadwater Slough as 1) few to no hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, less than 50% Chinook Salmon in the Northern Pikeminnow's diet may be more likely. Future work to quantify and identify juvenile salmonids in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model and to understand impacts to local populations.

@@

In additional to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon (both natural-origin and hatchery releases) are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that juvenile emigration and adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population. Predation on juveniles from any of the ESA-listed salmonid species is likely detrimental to their recovery.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Lastly, a more passive approach could be to provide cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover would need to be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor's Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from staff at Inter-Fluve. This manuscript benefited from reviews by Sean Gibbs, Mark Roes, and Ian Courter. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor's Office of Species Conservation.

>>>>

Predation on emigrating juveniles by piscivorous fishes is another important, and potentially under-estimated, source of salmonid mortality. In the lower mainstem Snake and Columbia rivers it is estimated that predation on out-migrating salmonids during peak emigration has a significant negative impact on the overall population and success of recovery efforts [@Friesen1999; @Tabor1993; @Beamesderfer1996]. Dams and reservoirs in the Columbia River are the primary locations associated with high rates of piscine predation on salmonids [@Petersen1994; @Ward1995]. There are generally two mechanisms that explain these high predation zones. First, migration (movement) rates of juvenile salmon are reduced during reservoir passage [@Venditti2000], thereby increasing the time migrating smolts are vulnerable to predation. Second, reservoirs and downstream tailraces associated with dams on the Snake and Columbia rivers create favorable slow-water habitat for predatory fishes known to consume juvenile salmonids, such as the Northern Pikeminnow *Ptychocheilus oregonsis*. Northern Pikeminnow tolerate and thrive in relatively warm, slow-water habitats [@Wydoski2003] and thus have benefited from dams on the Columbia River, becoming abundant predators of salmonid outmigrants [@Knutsen1999]. Northern Pikeminnow are estimated to consume \~8% (16.4 million) of the approximately 200 million juvenile salmonids emigrating through the lower Snake and Columbia Rivers, annually [@Beamesderfer1996].

@@

Chinook Salmon in the Upper Salmon River are stream-type and exhibit two distinct migration tactics; downstream rearing (DSR) and natal reach rearing (NRR) [@Copeland2014a]. The DSR migrants leave their natal area as subyearlings between June and November and typically overwinter in downstream, mainstem habitats until the following spring when they emigrate to the ocean as smolts. Alternatively, NRR migrants remain in their natal areas for approximately one year after emergence until emigration to the ocean as smolts. Diversity of migration tactics provides a mechanism for coping with adverse conditions in freshwater rearing and migration environments and buffers against catastrophic events, thereby increasing population resiliency [@Dodson2013]. Deadwater Slough represents an important habitat for Chinook Salmon in the Upper Salmon River as it supports rearing and migration of juveniles from all upstream populations including fall DSR and spring NRR migrants.

@@

Next, our abundance estimates assumed that capture and recapture events are random samples of the population. Angling methods often have a size selection bias, thereby limiting abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. The estimators also assumed equal catchability of individual fish between sampling events; two fish were recaptured during multiple days within a survey, indicating that marked fish were continually susceptible to angling. It is also possible that marked and released fish may not have immediately mixed thoroughly back into the population. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may have led to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. As an example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate. This would additionally account for marked and released fish being "hook shy" for a period of time. The adjusted delayed-mixing Schnabel estimator may account for these biases; however, the magnitude of bias is unknown.

@@

Our bioenergetics approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999]. Other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration, and so generalizing energy densities among salmonids may not be problematic. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during fall months when DSR Chinook Salmon are the dominant prey species available within the Salmon River and Deadwater Slough as 1) few to no hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, less than 50% Chinook Salmon in the Northern Pikeminnow's diet may be more likely. Future work to quantify and identify juvenile salmonids in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model and to understand impacts to local populations.

@@

In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon (both natural-origin and hatchery releases) are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that juvenile emigration and adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide for recreational fishing opportunities in the Upper Salmon River. Consequently, reducing predation mortality in Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population. Predation on juveniles from any of the ESA-listed salmonid species is likely detrimental to their recovery.

@@

Northern Pikeminnow bounty programs have proven successful at reducing population sizes and impacts on emigrating salmonids in the Columbia River [@Winther2020]. A local Northern Pikeminnow bounty program could provide monetary incentive for sportfisherman, in addition to boosting the local economy from lost revenues. Bounty programs could be conducted year-round or seasonally to coincide with peak juvenile outmigration(s). Each approach would require continual/annual effort and harvest to suppress the population size and as such, appropriate funding would be needed to support the program. Lastly, a more passive approach could be to improve cover or structure within Deadwater Slough that provides refugia for juvenile salmonids during their emigration. Cover would need to be appropriately sized to provide concealment to juvenile fishes while reducing access by larger fishes like Northern Pikeminnow. Adding cover could be a cost-effective approach; however, its potential effectiveness is unclear.

@@

The authors extend much appreciation to the many volunteers that assisted with field efforts including collaborators from the Bureau of Reclamation, Idaho Governor's Office of Species Conservation, Idaho Department of Fish and Game, and the Lemhi Regional Land Trust, among others. Special thanks to Braden Lott, Jared Barker, Brian Hamilton, Tulley Mackey, and Chelsea Welke for their help in the field. We further appreciate the administrative support and guidance from staff at Inter-Fluve. This manuscript benefited from reviews by Sean Gibbs, Mark Roes, and Ian Courter. Funding for this study was provided by the Bureau of Reclamation, Pacific Northwest Regional Office (contract No. 140R1021F0018) with assistance from the Idaho Governor's Office of Species Conservation.
